---
title: "EDA"
output: html_document
---

# Setup
## Load Packages
```{r}
library(nycflights13)
library(tidyverse)
library(ggplot2)
library(patchwork)
```

## Load Datasets
```{r}
# cast to dataframe
airlines_df = as.data.frame(airlines)
airports_df = as.data.frame(airports)
flights_df = as.data.frame(flights)
planes_df = as.data.frame(planes)
weather_df = as.data.frame(weather)
```


```{r}
airlines_df
head(airports_df)
head(flights_df)
head(planes_df)
head(weather_df)
```

```{r}
# creating PK for flights
flights_df$flight_num <- paste(flights_df$carrier, flights_df$flight, sep="-")
flights_df$key <- paste(flights_df$carrier, flights_df$flight, flights_df$time_hour, sep="-")

# also need to create PK for weather
weather_df$key <- paste(weather_df$origin, weather_df$time_hour, sep="-")
```

```{r}
# keeping track of which kind of variable each column is
airlines_categorical <- c("carrier", "name")

airports_categorical <- c("faa", "name", "dst", "tzone")
airports_ratio <- c("lat", "lon", "alt", "tz")

flights_categorical <- c("carrier", "tailnum", "origin", "dest", "flight", "flight_num") # may need to cast flight to another type since currently an int
flights_time <- c("year", "month", "day", "hour", "minute", "time_hour", "dep_time", "scheduled_dep_time", "arr_time", "scheduled_arr_time") # need to figure out how to deal with these... need to convert times to actual time
flights_ratio <- c("dep_delay", "arr_delay", "air_time", "distance")

planes_categorical <- c("tailnum", "type", "manufacturer", "model", "engine")
planes_ratio <- c("year", "engines", "seats", "speed") # idk if year really goes here

weather_categorical <- c("origin")
weather_time <- c("year", "month", "day", "hour", "time_hour") # need to figure out how to deal with these
weather_ratio <- c("temp", "dewp", "humid", "wind_dir", "wind_speed", "wind_gust", "precip", "pressure")
weather_ordinal <- c("visib")
```

```{r}
unique(flights_df$origin)
unique(flights_df$dest)
```

# Data Cleaning
```{r}
# check for NA
colSums(sapply(airports_df, is.na))
colSums(sapply(flights_df, is.na))
colSums(sapply(planes_df, is.na))
colSums(sapply(weather_df, is.na))
```
There are a few columns missing a lot of values. TODO: deal with this

# Summary Statistics
```{r}
# checking to see if each row can be identified uniquely
print("airlines")
length(airlines_df$carrier)
length(unique(airlines_df$carrier))

print("airports")
length(airports_df$faa)
length(unique(airports_df$faa))

print("flights")
length(flights_df$flight_num)
length(unique(flights_df$flight_num))
length(unique(flights_df$key))
length(unique(flights_df$tailnum))

print("planes")
length(planes_df$tailnum)
length(unique(planes_df$tailnum))

print("weather")
length(weather_df$key)
length(weather_df$key)
```
Length of dataframes:
airlines = 16
airports = 1,458
flights = 336,776
planes = 3,322
weather = 26,115

The number of unique tail numbers in flights do not match the number of unique tail numbers in planes, suggesting that there may be some missing records in the planes table. Additionally, the number of carrier + flight numbers in flights is much less than the table length, suggesting that this cannot be used as the primary key.

```{r}
max(flights_df$time_hour)
min(flights_df$time_hour)
```
## Numerical Summaries
```{r}
print("-----flights------")
summary(flights%>%select(all_of(flights_ratio)))
print("-----planes------")
summary(planes%>%select(all_of(planes_ratio)))
print("-----weather------")
summary(weather%>%select(all_of(weather_ratio)))
```

# Initial Visualizations

## Repeated Barplots
```{r, fig.width=16, fig.height=12}
barplot_grid <- function(df, cat_vars, color, n_col){
  cat_plots <- lapply(cat_vars, function(var) {
    ggplot(df, aes_string(x = var)) +
      geom_bar(fill = color) +
      labs(title = paste("Frequency of", var), x = var, y = "Relative Frequency") +
      theme_minimal()
  })
  
  return(wrap_plots(cat_plots, ncol = n_col))
}

barplot_grid(airports_df, airports_categorical, "steelblue", 3)
barplot_grid(flights_df, flights_categorical, "forestgreen", 3)
barplot_grid(planes_df, planes_categorical, "coral", 3)
```
The flights are relatively evenly distributed across the three airports, although EWR has the most flights and LGA has the least. The number of flights from each carrier differs, although some carriers use certain airports more often (e.g. UA and EV from EWR, 9E and B6 from JFK, MQ from LGA).

The number of flights by each plane and to each destination varies a lot.

## Repeated Boxplots
```{r, fig.width=10, fig.height=6}
boxplot_grid <- function(df, num_vars, color, n_col){
  box_plots <- lapply(num_vars, function(var) {
    ggplot(df, aes_string(x = var, y = "1")) +
      geom_boxplot(fill = color) +
      labs(title = paste("Boxplot of", var), x = NULL, y = var) +
      theme_minimal()
  })
  return(wrap_plots(box_plots, ncol = n_col))
}

boxplot_grid(airports_df, airports_ratio, "steelblue", 2)
boxplot_grid(flights_df, flights_ratio, "forestgreen", 2)
boxplot_grid(planes_df, planes_ratio, "coral", 2)
boxplot_grid(weather_df, weather_ratio, "darkorchid", 3)
```

## Repeated Histograms
```{r}
hist_grid <- function(df, num_vars, color, n_col){
  hist_plots <- lapply(num_vars, function(var) {
    ggplot(df, aes_string(x = var)) +
      geom_histogram(fill = color, bins = 10) +
      labs(title = paste("Histogram of", var), x = NULL, y = var) +
      theme_minimal()
  })
  return(wrap_plots(hist_plots, ncol = n_col))
}

hist_grid(airports_df, airports_ratio, "steelblue", 2)
hist_grid(flights_df, flights_ratio, "forestgreen", 2)
hist_grid(planes_df, planes_ratio, "coral", 2)
hist_grid(weather_df, weather_ratio, "darkorchid", 3)
```
Departure delay and arrival delay are skewed

## Faceted Plots
May want to pick a specific origin, although it doesn't look like there's a meaningful difference
```{r}
ggplot(data=flights_df, aes(x=carrier, fill=origin)) + geom_bar()
ggplot(data=flights_df, aes(x=log(dep_delay), fill=origin)) + geom_boxplot()
```

# Thoughts
Important response variables: arrival delay and departure delay (is there a meaningful difference between these? are they related?)
New variables:
direction (use lat and lon -- maybe slope between?)
could also combine and see what most important predictors are using dimensionality reduction technique

# Create New Aggregations
- combine sources
- by tail number (gap between flights; number of previous flights)
- tail number + date
- by date
- by origin and destination
- airline?